---
title: "Post-Count QC"
author: "CCBR RNA-seq pipeline"
output: html_document
params:
  folder: result
  sampleinfo: sampletable.txt
  data: RawCountFile_genes_filtered.txt
  projectId: "<projectId>"
  projectDesc: "<desc>"
---

```{r, include=FALSE, warning=FALSE, message=FALSE}
countMethod <- "subread"
if(grepl("rsem",params$data, ignore.case=TRUE)){
  countMethod <- "rsem"
}
projectId<-params$projectId
projectDesc<-params$projectDesc
dateandtime<-format(Sys.time(), "%a %b %d %Y - %X")
```
### **Project:**
####    *`r projectId`*
### **Description:** 
####    *`r projectDesc`*
###  **Count method:** 
####    *`r countMethod`* 
### **Report generated:** 
####    *`r dateandtime`* 


```{r setup, echo=FALSE, warning=FALSE,message=FALSE,fig.align='center'}

suppressMessages(library(rgl))
suppressMessages(library(knitr))
suppressMessages(library('edgeR'))
suppressMessages(library('statmod'))
suppressMessages(library('RColorBrewer')) 
suppressMessages(library('gplots'))
suppressMessages(library('reshape')) 
suppressMessages(library('ggplot2'))
suppressMessages(library('ggfortify'))
suppressMessages(library(ggdendro))
suppressMessages(library(amap))
suppressMessages(library(DT))
suppressMessages(library(plotly))
suppressMessages(library('geneplotter'))
suppressMessages(library('DESeq2'))

knit_hooks$set(rgl = function(before, options, envir) {
  if (!before) {
    ## after a chunk has been evaluated
    if (rgl.cur() == 0) return()  # no active device
    name = paste(options$fig.path, options$label, sep = '')
    rgl.snapshot(paste(name, '.png', sep = ''), fmt = 'png')
    return(paste('\\includegraphics{', name, '}\n', sep = ''))
  }
})

knit_hooks$set(webgl = hook_webgl)
```

```{r, echo=FALSE, warning=FALSE,message=FALSE}

## grab args & normalization

DIR <- params$folder
FILE1 <- params$sampleinfo
FILE2 <- params$data

# ANNOTATE <- args[4]

#
## setwd(DIR)
# read files
sampleinfo=read.delim(FILE1)
x = read.delim(FILE2,row.names=1)
colnames(x)=as.character(sampleinfo[,4])
sampleFiles=as.character(sampleinfo[,2])
## read annotation file
## ann=read.delim(ANNOTATE)
# DGElist object --------------------------------------------------------------
condition = as.factor(sampleinfo$condition)
y = DGEList(counts=x,group=condition)
## Normalization TMM ------------------------------------------------------------
## method = =c("TMM","RLE","upperquartile","none")
y <- calcNormFactors(y,method="TMM")
# y$samples

Group <- factor(sampleinfo$condition)
if ( length(levels(Group)) == 1 ) {
  design=~1
  v1 <- voom(as.matrix(x),design=NULL,plot=FALSE,normalize="quantile")
} else {
  design=model.matrix(~0+Group)
  v1 <- voom(as.matrix(x),design,plot=FALSE,normalize="quantile")
}
ddsHTSeq<-DESeqDataSetFromMatrix(countData=x,colData=sampleinfo, design=design)
dds<-DESeq(ddsHTSeq)
dds.ndata=as.data.frame(counts(dds,normalized=TRUE))


## estimating common and tagwise dispersions -----------------------------------------
y <- estimateCommonDisp(y)
y <- estimateTagwiseDisp(y) #default trend: moveingave

ylog2=cpm(y,log=TRUE,normalized.lib.sizes=TRUE,prior.count=0.5) # prior count like avelogcpm
rawlog2= cpm(y,log=TRUE,normalized.lib.sizes=FALSE,prior.count=0.5)
#ddslog2= cpm(dds.ndata,log=TRUE,normalized.lib.sizes=FALSE,prior.count=0.5) 

rld <- rlogTransformation(dds, blind=TRUE)
rldm=assay(rld)
colnames(rldm)=colnames(x)

## save it
```

## **Count Distribution** {.tabset}

### Before Normalization

```{r, echo=FALSE, warning=FALSE,message=FALSE}
beforehist <- ggplotly(ggplot(melt(as.data.frame(rawlog2))) + geom_line(stat="density", aes(x = value,colour = variable)) + labs(x = NULL) + theme(legend.position='right') + scale_x_log10())
beforehist
```

### TMM

```{r, echo=FALSE, warning=FALSE,message=FALSE}
tmmhist <- ggplotly(ggplot(melt(as.data.frame(ylog2))) + geom_line(stat="density", aes(x = value,colour = variable)) + labs(x = NULL) + theme(legend.position='right') + scale_x_log10())
tmmhist
```

### DESeq2

```{r, echo=FALSE, warning=FALSE,message=FALSE}
deshist <- ggplotly(ggplot(melt(as.data.frame(rldm))) + geom_line(stat="density", aes(x = value,colour = variable)) + labs(x = NULL) + theme(legend.position='right') + scale_x_log10())
deshist
```

### Limma

```{r, echo=FALSE, warning=FALSE,message=FALSE}
limmahist <- ggplotly(ggplot(melt(as.data.frame(v1$E))) + geom_line(stat="density", aes(x = value,colour = variable)) + labs(x = NULL) + theme(legend.position='right') + scale_x_log10())
limmahist
```

## **PCA Plots** {.tabset}

### Before Normalization

```{r, echo=FALSE,webgl=TRUE,message=FALSE, warning=FALSE}

## PCA for before norm

before.edf=rawlog2
before.tedf= t(before.edf)
before.pca=prcomp(before.tedf,scale.=T)
before.tedf1 = data.frame(before.tedf)
Phenotype=sampleinfo$condition
cell_rep=sampleinfo$label
before.tedf1$group = as.factor(Phenotype)
before.pc1 = round(before.pca$sdev[1]^2/sum(before.pca$sdev^2)*100,2)
before.pc2 = round(before.pca$sdev[2]^2/sum(before.pca$sdev^2)*100,2)
before.pc3 = round(before.pca$sdev[3]^2/sum(before.pca$sdev^2)*100,2)

pcafactor = as.factor(sampleinfo$condition)

library(RColorBrewer)

col <- brewer.pal(nlevels(pcafactor), "Paired")

p <- plot_ly(as.data.frame(before.pca$x[,1:3]), x = ~PC1, y = ~PC2, z = ~PC3, color = pcafactor, colors = col, hoverinfo="text",
             hovertext = ~sampleinfo$label) %>%
  add_markers() %>%
  layout(title = "Before Normalization PCA plot", 
         scene = list(xaxis = list(title = paste0("PC1 (",before.pc1,"%)")),
                      yaxis = list(title = paste0("PC2 (",before.pc2,"%)")),
                      zaxis = list(title = paste0("PC3 (",before.pc3,"%)"))))

p

# plot(before.pca,type="lines")  #Decide how many PC's are relevant for plotting
  #before.pca$x[,1:3]  #look at first 3 PC's

#plot3d(before.pca$x[,1:3],col = as.integer(before.tedf1$group),type="s",size=2,main="PCA before normalization",xlab=paste0("PC1 (",before.pc1,"%)"),ylab=paste0("PC2 (",before.pc2,"%)"),zlab=paste0("PC3 (",before.pc3,"%)"))
#group.v<-as.vector(cell_rep)
#text3d(before.pca$x, before.pca$y, before.pca$z, group.v, cex=1.0, adj = 1.2)
#legend3d("topright", legend = levels(sampleinfo$condition), pch = 16, col = as.numeric(as.factor(levels(sampleinfo$condition))), cex=0.5)
#rgl.postscript("pca3d_raw.pdf","pdf")
#rgl.snapshot("pca3d_raw.png","png")

```

### TMM

```{r, echo=FALSE,webgl=TRUE,message=FALSE, warning=FALSE}

## PCA for edgeR

edgeR.edf=ylog2
edgeR.tedf= t(edgeR.edf)
edgeR.pca=prcomp(edgeR.tedf,scale.=T)
edgeR.tedf1 = data.frame(edgeR.tedf)
Phenotype=sampleinfo$condition
cell_rep=sampleinfo$label
edgeR.tedf1$group = as.factor(Phenotype)
edgeR.pc1 = round(edgeR.pca$sdev[1]^2/sum(edgeR.pca$sdev^2)*100,2)
edgeR.pc2 = round(edgeR.pca$sdev[2]^2/sum(edgeR.pca$sdev^2)*100,2)
edgeR.pc3 = round(edgeR.pca$sdev[3]^2/sum(edgeR.pca$sdev^2)*100,2)

pcafactor = as.factor(sampleinfo$condition)

library(RColorBrewer)

col <- brewer.pal(nlevels(pcafactor), "Paired")

p <- plot_ly(as.data.frame(edgeR.pca$x[,1:3]), x = ~PC1, y = ~PC2, z = ~PC3, color = pcafactor, colors = col, hoverinfo="text",
             hovertext = ~sampleinfo$label) %>%
  add_markers() %>%
  layout(title = "edgeR PCA plot", 
         scene = list(xaxis = list(title = paste0("PC1 (",edgeR.pc1,"%)")),
                      yaxis = list(title = paste0("PC2 (",edgeR.pc2,"%)")),
                      zaxis = list(title = paste0("PC3 (",edgeR.pc3,"%)"))))

p

# plot(edgeR.pca,type="lines")  #Decide how many PC's are relevant for plotting
  #edgeR.pca$x[,1:3]  #look at first 3 PC's

#plot3d(edgeR.pca$x[,1:3],col = as.integer(edgeR.tedf1$group),type="s",size=2,main="PCA after TMM normalization",xlab=paste0("PC1 (",edgeR.pc1,"%)"),ylab=paste0("PC2 (",edgeR.pc2,"%)"),zlab=paste0("PC3 (",edgeR.pc3,"%)"))
#group.v<-as.vector(cell_rep)
#text3d(edgeR.pca$x, edgeR.pca$y, edgeR.pca$z, group.v, cex=1.0, adj = 1.2)
#legend3d("topright", legend = levels(sampleinfo$condition), pch = 16, col = as.numeric(as.factor(levels(sampleinfo$condition))), cex=0.5)
#rgl.postscript("pca3d_edgeR.pdf","pdf")
#rgl.snapshot("pca3d_edgeR.png","png")

```

### DESeq2

```{r, echo=FALSE,webgl=TRUE,message=FALSE, warning=FALSE}

## PCA for deseq2

#rld <- rlogTransformation(dds, blind=TRUE)
#rldm=assay(rld)
#colnames(rldm)=colnames(x)
deseq2.edf=as.matrix(rldm)
#deseq2.edf=ddslog2
deseq2.tedf= t(deseq2.edf)
deseq2.tedf=deseq2.tedf[,apply(deseq2.tedf,2,var)!= 0]
deseq2.pca=prcomp(deseq2.tedf,scale.=T)
deseq2.tedf1 = data.frame(deseq2.tedf)
Phenotype=sampleinfo$condition
cell_rep=sampleinfo$label
deseq2.tedf1$group = as.factor(Phenotype)
deseq2.pc1 = round(deseq2.pca$sdev[1]^2/sum(deseq2.pca$sdev^2)*100,2)
deseq2.pc2 = round(deseq2.pca$sdev[2]^2/sum(deseq2.pca$sdev^2)*100,2)
deseq2.pc3 = round(deseq2.pca$sdev[3]^2/sum(deseq2.pca$sdev^2)*100,2)

pcafactor = as.factor(sampleinfo$condition)

library(RColorBrewer)

col <- brewer.pal(nlevels(pcafactor), "Paired")

p <- plot_ly(as.data.frame(deseq2.pca$x[,1:3]), x = ~PC1, y = ~PC2, z = ~PC3, color = pcafactor, colors = col, hoverinfo="text",
             hovertext = ~sampleinfo$label) %>%
  add_markers() %>%
  layout(title = "DESeq2 PCA plot", 
         scene = list(xaxis = list(title = paste0("PC1 (",deseq2.pc1,"%)")),
                      yaxis = list(title = paste0("PC2 (",deseq2.pc2,"%)")),
                      zaxis = list(title = paste0("PC3 (",deseq2.pc3,"%)"))))

p

#plot3d(deseq2.pca$x[,1:3],col = as.integer(deseq2.tedf1$group),type="s",size=2,main="PCA after DESeq2 normalization",xlab=paste0("PC1 (",deseq2.pc1,"%)"),ylab=paste0("PC2 (",deseq2.pc2,"%)"),zlab=paste0("PC3 (",deseq2.pc3,"%)"))
#group.v<-as.vector(cell_rep)
#text3d(deseq2.pca$x, deseq2.pca$y, deseq2.pca$z, group.v, cex=1.0, adj = 1.2)
#legend3d("topright", legend = levels(sampleinfo$condition), pch = 16, col = as.numeric(as.factor(levels(sampleinfo$condition))), cex=0.5)
#rgl.postscript("pca3d_deseq2.pdf","pdf")
#rgl.snapshot("pca3d_deseq2.png","png")

```

### Limma

```{r, echo=FALSE,webgl=TRUE,message=FALSE, warning=FALSE}

## PCA for Limma

limma.edf=as.matrix(v1$E)
limma.tedf= t(limma.edf)
limma.tedf=limma.tedf[,apply(limma.tedf,2,var)!= 0]
limma.pca=prcomp(limma.tedf,scale.=T)
limma.tedf1 = data.frame(limma.tedf)
Phenotype=sampleinfo$condition
cell_rep=sampleinfo$label
limma.tedf1$group = as.factor(Phenotype)
limma.pc1 = round(limma.pca$sdev[1]^2/sum(limma.pca$sdev^2)*100,2)
limma.pc2 = round(limma.pca$sdev[2]^2/sum(limma.pca$sdev^2)*100,2)
limma.pc3 = round(limma.pca$sdev[3]^2/sum(limma.pca$sdev^2)*100,2)

pcafactor = as.factor(sampleinfo$condition)

library(RColorBrewer)

col <- brewer.pal(nlevels(pcafactor), "Paired")

p <- plot_ly(as.data.frame(limma.pca$x[,1:3]), x = ~PC1, y = ~PC2, z = ~PC3, color = pcafactor, colors = col, hoverinfo="text",
             hovertext = ~sampleinfo$label) %>%
  add_markers() %>%
  layout(title = "Limma PCA plot", 
         scene = list(xaxis = list(title = paste0("PC1 (",limma.pc1,"%)")),
                      yaxis = list(title = paste0("PC2 (",limma.pc2,"%)")),
                      zaxis = list(title = paste0("PC3 (",limma.pc3,"%)"))))

p


#plot3d(limma.pca$x[,1:3],col = as.integer(limma.tedf1$group),type="s",size=2,main="PCA after Limma normalization",xlab=paste0("PC1 (",limma.pc1,"%)"),ylab=paste0("PC2 (",limma.pc2,"%)"),zlab=paste0("PC3 (",limma.pc3,"%)"))
#group.v<-as.vector(cell_rep)
#text3d(limma.pca$x, limma.pca$y, limma.pca$z, group.v, cex=1.0, adj = 1.2)
#legend3d("topright", legend = levels(sampleinfo$condition), pch = 16, col = as.numeric(as.factor(levels(sampleinfo$condition))), cex=0.5)
#rgl.postscript("pca3d_limma.pdf","pdf")
#rgl.snapshot("pca3d_limma.png","png")

```

```{r, echo=FALSE,message=FALSE,warning=FALSE}
before.dfm <- melt(as.data.frame(rawlog2))
edgeR.dfm <- melt(as.data.frame(ylog2))
deseq2.dfm <- melt(as.data.frame(rldm))
limma.dfm <- melt(as.data.frame(v1$E))
```

## **Boxplots** {.tabset}

### Before Normalization

```{r, echo=FALSE,message=FALSE,fig.show='hold',fig.align='center', warning=FALSE}
par(mar=c(par("mar")[1]+5,par("mar")[-1]))
boxplot(value~variable,las=2,data=before.dfm,main="Before normalization",ylab="Counts",col=as.numeric(as.factor(sampleinfo$condition)))
```

### TMM

```{r, echo=FALSE,message=FALSE,fig.show='hold',fig.align='center', warning=FALSE}
par(mar=c(par("mar")[1]+5,par("mar")[-1]))
boxplot(value~variable,las=2,data=edgeR.dfm,main="TMM",ylab="Counts",col=as.numeric(as.factor(sampleinfo$condition)))
```

### DESeq2

```{r, echo=FALSE,message=FALSE,fig.show='hold',fig.align='center', warning=FALSE}
par(mar=c(par("mar")[1]+5,par("mar")[-1]))
boxplot(value~variable,las=2,data=deseq2.dfm,main="DESeq2",ylab="Counts",col=as.numeric(as.factor(sampleinfo$condition)))
```

### Limma

```{r, echo=FALSE,message=FALSE,fig.show='hold',fig.align='center', warning=FALSE}
par(mar=c(par("mar")[1]+5,par("mar")[-1]))
boxplot(value~variable,las=2,data=limma.dfm,main="Limma",ylab="Counts",col=as.numeric(as.factor(sampleinfo$condition)))
```

## **Similarity Heatmaps Using Pearson - Complete Linkage** {.tabset}

```{r, echo=FALSE,message=FALSE, warning=FALSE}
bothdendro = "both"
hmapsize = floor(length(sampleinfo$label)/20) + 8
if(hmapsize>20){
hmapsize = 15
bothdendro = "none"
}

create_heatmap <- function(data){  
    dd_col <- as.dendrogram(hclust(dist(data)))
    dd_row <- as.dendrogram(hclust(dist(t(data))))
    dendro_1 <- dendro_data(dd_col)
    dendro_2 <- dendro_data(dd_row)
    
    hmcol <- colorRampPalette(c("black","red","yellow","white"),space="rgb")(100)
    
    ggdend <- function(df) {
      ggplot() +
        geom_segment(data = df, aes(x=x, y=y, xend=xend, yend=yend)) +
        labs(x = "", y = "") + theme_minimal() +
        theme(axis.text = element_blank(), axis.ticks = element_blank(),
              panel.grid = element_blank())
    }
    
    dendro_columns <- ggdend(dendro_1$segments)
    dendro_rows <- ggdend(dendro_2$segments) + coord_flip()
    
    melt_mat <- melt(data)
    hmap <- ggplot(data = melt_mat, aes(x=X1, y=X2, fill=value)) + 
           geom_tile() +
           scale_fill_gradientn(colours = rev(hmcol), limit = c(0,max(data)), space = "Lab", name="Correlation") +
           theme_minimal()+ 
           theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 10, hjust = 1),
                 legend.justification = c(1, 0),
                 legend.position = c(0.6, 0.7),
                 legend.direction = "horizontal",
                 axis.title.x = element_blank(),
                 axis.title.y = element_blank()) +
           coord_fixed()
    
    eaxis <- list(showticklabels = FALSE, showgrid = FALSE, zeroline = FALSE)
    p_empty <- plot_ly() %>% layout(margin = list(l = 200), xaxis = eaxis,yaxis = eaxis)
    
    sply <- subplot(dendro_columns, p_empty, hmap, dendro_rows, nrows = 2)
    
    sply <- layout(sply,
                   yaxis = list(domain=c(0.47, 1)),
                   xaxis = list(domain=c(0, 0.5)),
                   xaxis3 = list(domain=c(0, 0.5)),
                   xaxis4 = list(domain=c(0.5, 1)),
                   margin = list(l = 150, r = 0, b = 50,t = 0))
    
    return(sply)
}

```

### Before Normalization

```{r, echo=FALSE,message=FALSE,fig.show='hold',fig.align='center', warning=FALSE, fig.width=hmapsize, fig.height=hmapsize}

hmcol <- colorRampPalette(c("black","red","yellow","white"),space="rgb")(100)
before.distrawlog2=amap::Dist(t(rawlog2),method="pearson")
before.mat = as.matrix(before.distrawlog2)

heatmap.2(before.mat, trace="none", col = rev(hmcol), labCol=FALSE, Rowv=TRUE, Colv= TRUE, dendrogram = bothdendro, colRow=as.numeric(as.factor(sampleinfo$condition)), margin=c(16, 16), main="Before normalization")
```

### TMM

```{r, echo=FALSE,message=FALSE,fig.show='hold',fig.align='center', warning=FALSE, fig.width=hmapsize, fig.height=hmapsize}

edgeR.distylog2=amap::Dist(t(ylog2),method="pearson")
edgeR.mat = as.matrix(edgeR.distylog2)
heatmap.2(edgeR.mat, trace="none", col = rev(hmcol), labCol=FALSE, Rowv=TRUE, Colv= TRUE, dendrogram = bothdendro, colRow=as.numeric(as.factor(sampleinfo$condition)), margin=c(16, 16), main="TMM")
```

### DESeq2

```{r, echo=FALSE,message=FALSE,fig.show='hold',fig.align='center', warning=FALSE, fig.width=hmapsize, fig.height=hmapsize}

deseq2.dists <- amap::Dist(t(rldm),method="pearson")
deseq2.mat <- as.matrix(deseq2.dists)
heatmap.2(deseq2.mat, trace="none", col = rev(hmcol), labCol=FALSE, Rowv=TRUE, Colv= TRUE, dendrogram = bothdendro, colRow=as.numeric(as.factor(sampleinfo$condition)), margin=c(16, 16), main="DESeq2")
```

### Limma

```{r, echo=FALSE,message=FALSE,fig.show='hold',fig.align='center', warning=FALSE, fig.width=hmapsize, fig.height=hmapsize}

limma.dists <- amap::Dist(t(limma.edf),method="pearson")
limma.mat <- as.matrix(limma.dists)
heatmap.2(limma.mat, trace="none", col = rev(hmcol), labCol=FALSE, Rowv=TRUE, Colv= TRUE, dendrogram = bothdendro, colRow=as.numeric(as.factor(sampleinfo$condition)), margin=c(16, 16), main="Limma")
```


## **MD Plots** (sample by sample mean vs fold change of expression)

```{r, echo=FALSE,message=FALSE, warning=FALSE}
mdchunkheight <- 3.3*length(sampleinfo$label)
mdflag <- TRUE

if(mdchunkheight>100){
 mdchunkheight = 8
 mdflag <- FALSE
}
```

```{r, echo=FALSE,message=FALSE,fig.show='hold',fig.align='center', warning=FALSE, fig.height=mdchunkheight}
if(mdflag){
  par(mfcol=c(length(sampleinfo$label),4))
  for(i in 1:length(sampleinfo$label)){
    plotMD(rawlog2,column=i, main=paste0("Raw ",sampleinfo$label[i]), xlim=c(-5,15), ylim=c(-15,15))
    abline(h=0, col="red", lty=2, lwd=2)
  }
  for(i in 1:length(sampleinfo$label)){
    plotMD(ylog2,column=i, main=paste0("TMM ",sampleinfo$label[i]), xlim=c(-5,15), ylim=c(-15,15))
    abline(h=0, col="red", lty=2, lwd=2)
  }
  for(i in 1:length(sampleinfo$label)){
    plotMD(rldm,column=i, main=paste0("DESeq2 ",sampleinfo$label[i]), xlim=c(-5,15), ylim=c(-15,15))
    abline(h=0, col="red", lty=2, lwd=2)
  }
  for(i in 1:length(sampleinfo$label)){
    plotMD(v1$E,column=i, main=paste0("Limma ",sampleinfo$label[i]), xlim=c(-5,15), ylim=c(-15,15))
    abline(h=0, col="red", lty=2, lwd=2)
  }
}
```
